---
title: "Static plots - Assignments"
output:
  word_document: default
  html_document: default
---

### References and datasets

-   Original paper: Abbas et al. (2019) *Ecological Indicators* (<https://doi.org/10.1016/j.ecolind.2019.105524>)
-   StatQuest: PCA and other videos (<https://statquest.org/>)
-   Also check the lecture slides for L02-L07

# 1. Student ID and Name

## [Updated: You do not need to write your student ID and name]{style="color:red"}

# 2. Generative AI declaration

Students may use generative AI tools during assignment, for example, to check and improve code, provided that the content is original and reflects the student's own contributions. If generative AI is used, students must declare its use in the box below. An example statement is: “*During the assignment, I used [XXXX] to [XXXX]. I subsequently reviewed and edited the code and take(s) full responsibility for the final version of the code.”*

::: textbox
During the assignment, I used Claude to help structure the sequence of my parameters for visualising the results of the PCA and NDMS analysis. I subsequently reviewed and edited the code and take full responsibility for the final version of the code.
:::

# 3. Assignments

## Github: Create a Github account

-   Create your Github account.
-   Install Git on your computer as well (the latest version = 2.53.0).
-   Not mandatory, but please read the advice about usernames in the lecture.

## Github: Create your first repository

-   Repository name must be “DASC3240_Assignment1.”
-   Make the visibility “public” so I can check it.
-   The repository must have a README and a License file.
-   Include a screenshot of your shell window showing the version of Git on your computer.

## Tree community data in Hong Kong

Go to Canvas and download `Abbas_2019_TableS4_ed1.csv` and `Abbas_2019_TableS4_ed2.csv`.

-   `Abbas_2019_TableS4_ed1.csv` includes the plot information (metadata).
-   `Abbas_2019_TableS4_ed2.csv` includes dominant tree species data in the 28 forest plots.

1.  Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age. You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.

2.  Visualize the dominant tree species compositions using dimension reduction techniques. For visualization, you must use different colors for different forest age groups.

    2-1. Perform PCA to visualize the tree species composition.\
    2-2. Perform NMDS to visualize the tree species composition.\

# 3. Marking criteria and important notes

[**Please check the lecture PDF "L08_Assignment-I.pdf" provided on Canvas.**]{style="color:darkred"}

# 4. Your answers

### 1. Calculate the total abundance of the dominant tree species in each forest plot, and visualize the relationship between the total abundance and forest age.

-   You should use `AGE_MEDIAN` column in the `Abbas_2019_TableS4_ed1.csv` file as a forest age index.
-   You can calculate **the sum of each column** in `Abbas_2019_TableS4_ed2.csv` to calculate the total abundance of the dominant tree species in each plot.
-   You will combine the data from two dataframes, so please check if the column names of `Abbas_2019_TableS4_ed2.csv` and the row names (indices) of `Abbas_2019_TableS4_ed1.csv` match (by using R script, not by visual inspection!).
-   To visualize the relationship, barplot is **NOT** recommended.

```{r}
# Loading the tidyverse library, which includes ggplot2 for plotting
library(tidyverse)

#Loading the two CSV files

# row.names = 1 uses the first column (PLOT) as the row names
d_meta <- read.csv("data/Abbas_2019_TableS4_ed1.csv", row.names = 1)

# d_tree: tree species abundance data
# row.names = 1 uses the first column (tree_species) as row names
d_tree <- read.csv("data/Abbas_2019_TableS4_ed2.csv", row.names = 1)

# Checking that the column names of d_tree match the row names of d_meta 
# all() returns TRUE only when every comparison is TRUE
all(colnames(d_tree) == rownames(d_meta))

# Calculateing the total abundance of all tree species in each forest plot
# colSums() sums each column i.e. adds up all species within each plot
total_abundance <- colSums(d_tree)

# Combining total abundance and forest age into one data frame for plotting
# AGE_MEDIAN is used as the forest age index
d_combined <- data.frame(total_abundance = total_abundance,
                         AGE_MEDIAN      = d_meta$AGE_MEDIAN)


# Multiple forest plots sharing the same AGE_MEDIAN value -> a boxplot grouped by age class is therefore a good way to show the distribution.

# factor(AGE_MEDIAN) treats the numeric age values as discrete groups on the x-axis.

# Overlaying a jitter plot on top of the boxplot
# geom_boxplot(outliers = FALSE): draws the boxplot but hides the default outlierpoints. This is important because geom_jitter() will already show all points, so leaving outliers = TRUE would draw those points twice.

d_combined %>%
  ggplot(aes(x = factor(AGE_MEDIAN), y = total_abundance)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(height = 0, width = 0.2, size = 2, alpha = 0.8) +
  labs(title = "Total tree abundance vs. forest age",
       x = "Median forest age (yr)",
       y = "Total tree abundance (ind.)") +
  NULL
```

### 2. Visualize the dominant tree species compositions using PCA and NMDS.

There are a number of points you should consider:

-   For visualization, you must use different colors for different forest age groups. Use of different marker styles for different groups is also a good idea.
-   [**Set random seeds if your analysis includes random process(es)**]{style="color:darkred"}.
-   Before applying dimension reduction functions (e.g., `prcomp()`), please make sure that [**rows represent samples (= forest plots) and columns represent tree species**]{style="color:darkred"}. You have to transpose the data frame (In `Abbas_2019_TableS4_ed2.csv`, columns represent samples and rows represent tree species).
-   Usually we should consider whether we must perform the data preprocessing before applying dimension reduction techniques (e.g., standardization). [**In this assignment, please convert your data into compositional data using the following code**]{style="color:darkred"} `d_tree_rel <- apply(d_tree, 2, function(x) x/sum(x))` (`d_tree` represents the name of your data.frame. You should change it depending on your R object name). This means that, in this assignment, we are interested in tree community compositions, not abundances.

```{r}
# Result: each column (plot) now sums to 1 since here we are looking at proportions
d_tree_rel <- apply(d_tree, 2, function(x) x/sum(x))

# Transpose so that rows = forest plots and columns = tree species
# This orientation is required by both prcomp() and metaMDS()
d_tree_t <- t(d_tree_rel)
```

#### 2-1. Perform PCA to visualize the tree species composition.

After PCA, please visualize the result.

```{r}
# prcomp() runs PCA
# center = FALSE, scale = FALSE: no further centring or scaling is needed 
pca_result <- prcomp(d_tree_t, center = FALSE, scale = FALSE)

# Summary of how much variance each PC captures
summary(pca_result)

# Calculating the proportion of variance explained by each PC
pca_vars <- pca_result$sdev^2
prop_vars <- pca_vars / sum(pca_vars)


# Visualising PCA results

# Extracting the PCA scores: the coordinates of each plot in PCA space
pca_df <- as.data.frame(pca_result$x)

# Adding the forest age information so we can colour the points
# AGE_MEDIAN (values 7, 20, 39, 61, 70) as the grouping variable, wrapped in factor() so ggplot treats it as discrete categories
pca_df$AGE_MEDIAN <- d_meta$AGE_MEDIAN

# Plot PC1 vs PC2
# colour and shape both vary by AGE_MEDIAN so the groups are easy to tell apart
pca_df %>%
  ggplot(aes(x = PC1, y = PC2, 
             color = factor(AGE_MEDIAN), 
             shape = factor(AGE_MEDIAN))) +
  geom_point(size = 3) +
  labs(title = "PCA of tree species composition",
       x = paste0("PC1 (", round(prop_vars[1] * 100, 1), "%)"),
       y = paste0("PC2 (", round(prop_vars[2] * 100, 1), "%)"),
       color = "AGE_MEDIAN",
       shape = "AGE_MEDIAN") +
  NULL

```

#### 2-2. Perform NMDS to visualize the tree species composition.

After NMDS, please visualize the result. You can use Bray-Curtis dissimilarity to calculate the dissimilarity among plots.

```{r}
# Loading the vegan package, which provides the metaMDS() function for NMDS
library(vegan)

# Setting a random seed BEFORE running NMDS
# NMDS starts from a random configuration, so setting a seed makes the result reproducible so we get the same plot every time we run this code
set.seed(1234)

# metaMDS() runs NDMS
# d_tree_t: our transposed compositional data (rows = plots, columns = species)
# distance = "bray": Bray-Curtis dissimilarity
# k = 2: reduce to 2 dimensions so we can plot the result easily
nmds_res <- metaMDS(d_tree_t, distance = "bray", k = 2)



# Visualising NMDS results

# nmds_res$points gives the 2D coordinates of each forest plot
# AGE_MEDIAN from d_meta is used for colouring and shape, the same grouping variable used in the PCA plot above

nmds_res$points %>% as.data.frame %>%
  ggplot(aes(x = MDS1, y = MDS2,
             color = factor(d_meta$AGE_MEDIAN),
             shape = factor(d_meta$AGE_MEDIAN))) +
  geom_point(size = 3) +
  xlab("Axis 1") + ylab("Axis 2") +
  labs(title = "NMDS of tree species composition (Bray-Curtis)",
       color = "AGE_MEDIAN",
       shape = "AGE_MEDIAN") +
  NULL

```
